Mixed language processing increases cross-language phonetic transfer in Bengali-English bilinguals

### 2022-10-10 draft_BLC/blc_revision1_reviewer_response.txt

Manuscript BLC-21-0133.R1
Response to Reviewers

Dear Editors,

Thank you for the opportunity to submit a revised draft of our manuscript “Mixed language processing increases cross-language phonetic transfer in Bengali-English bilinguals” for publication in Bilingualism: Language and Cognition. We appreciate the time and effort from your end and that of the reviewer in providing feedback on our manuscript. We have incorporated the suggestions made by the reviewer. Please see below, for a point-by-point response to the reviewer's comments and concerns. All page numbers refer to the revised manuscript file uploaded in the portal.

@@

Brysbaert, M. (2019). How many participants do we have to Include in properly powered experiments? A tutorial of power analysis with reference tables [Review of How many participants do we have to Include in properly powered experiments? A tutorial of power analysis with reference tables]. Journal of Cognition, 2, 16.
Brysbaert, M. (2021). Power considerations in bilingualism research: Time to step up our game. Bilingualism: Language and Cognition, 24(5), 813–818.
Cumming, G. (2013). Understanding the New Statistics: Effect Sizes, Confidence Intervals, and Meta-Analysis. Routledge.
Lakens, D. (2022). Sample size justification. Collabra. Psychology, 8(1). https://doi.org/10.1525/collabra.33267

Authors' response (AR): Thank you for facilitating a productive discussion on this issue, and for pointing us to these references. We have removed the power analysis from the main manuscript. As per the editor's suggestion, we have included it as Supplementary Material. As suggested, we have acknowledged the small sample size and need for replication in the main text, on page ...

@@

The second comment of substance I have has to do with the procedure. I do not fully understand the procedure of the cued-picture naming task. Can you please explain it in greater detail and perhaps provide some sample trials? You could give the examples in the text and then also a schema in a Figure or something like that. To be honest, I don’t know whether the participants are reading out loud (and the cues are orthographic) or they are naming pictures (and then I don’t understand how language is cued). Please, just explain, give examples, and consider adding a schema in a figure. (“The picture of a fish was used in all {sic} trials…” ????)

AR: We have now added a schematic figure showing examples of a nonswitched and a switched trial (figure 4, page...). The trials proceeded as follows:
(i) see a word in Bengali/English orthography. This is the cue. It is NOT read aloud.
(ii)see a picture of a fish. Based on the language cue in the previous slide, name it aloud in either English (say "fish") or Bengali (say "matS").  
(iii) see an English word, read it aloud. This is the target. If the previous slide was named in English this counts as a nonswitch trial; if it was named in Bengali this counts as a switched trial. 
(iv) solve a distractor match problem before proceeding to the next trial

@@

The third comment of substance, which I hadn’t noticed before, has to do with acoustic analyses. On page 26, it is said that formant values were Lobanov-normalized. I don’t think this is possible in this study. I believe the authors did something akin to Lobanov-normalization but quite distinct from it. In Lobanov normalization, values are z-scored with reference to values of all the vowels in a speaker’s vowel system. In other words, for one to be able to do a Lobanov-normalization for a given vowel category (the tokens from a category) one MUST have tokens, collected from the same speaker, of all the other vowels in the speaker’s vowel system. In this case, the given vowel token is expressed with reference to its location in a speaker’s vowel space. This DOES allow to normalize across genders. This is a great normalization procedure, but one needs to have collected samples of all the vowels in the system, or at least of the corner vowels, which outline the limits of the system. I don’t think this is what the authors did here (because I don’t think they collected samples of all the vowels of English or at least the corner vowels). Their normalization method is akin to the Lobanov method in that they did z-scoring with all the vowels they had. However, if they didn’t have samples of all the vowels, this is not a true normalization method, and clearly not Lobanov’s normalization procedure. Please, explain better and then explain what this normalization method accomplished. (Since mixed-effects regression was to be used, and all factors are within-speaker, normalization was likely unnecessary.)

AR: Thanks for pointing this out; we agree that "Lobanov normalization" is not the appropriate term here, given the lack of all corner vowels. We have replaced this with the term "z-score normalization". The target was to represent all formant values on a uniform scale across speakers, removing speaker-specific value ranges, so that the relative positions of the two vowels and shift between conditions could be compared directly. Since the data had productions of the same two categories from all speakers, we felt comfortable that this would not affect the interpretability of the values. This was useful for first-pass visualizations to understand speakers' vowel systems (e.g. fig. 2 in the text, which shows that speakers vary greatly in their default realizations of the two IE vowels), and visualize gender differences without confounds from individual differences (e.g. figure 6). Since we were expecting high individual variability, we did this before running any further analyses. We agree that given the specific statistical tools used here, this was not essential to the statistical analyses. We re-ran the analysis with non-normalized values to verify this, and found identical results. We have included the code and output from both analyses in the OSF repository for the project, and added a footnote in the main text addressing this (pg. ...).

@@

A quick note about terminology. I am a bit uncomfortable with the authors’ use of the terms “unilingual” and “mixed.” In past research, “unilingual” tends to refer to entire recording sessions in which a single language was used, and “bilingual” refers to recording sessions in which two languages were used. I don’t object to the use of “mixed” here, but only to the use of “unilingual.” It seems that all recording sessions in this study were bilingual or mixed. The difference between the two condition has to do with whether there was a code- or language-switch, that is, whether the target English word was produced after another English word or after a Bengali word. Following the terminology used by others (see Olson’s articles, for instance), I believe you should use “switch” vs “nonswitch” contexts or conditions rather than “unilingual” vs “mixed”. If not, at least explain your use of terminology at the outset and why you use it differently from others.

AR: We have replaced the terminology throughout the text, and used "nonswitch" and "switch/mixed" to refer to the two conditions.

@@

Finally, I have a comment about punctuation. The authors, in many locations, use the n-dash (or minus sign) instead of the colon. One example: Page 24, line 51. “Language cue- a word…” I have seen this use of the n-dash throughout the manuscript. I don’t think this follows convention. I believe a colon should be used in most places where an n-dash is used. Please, consider, decide, and revise throughout if needed.

AR: We have revised these.

The following are minor comments:

AR: By "atomic", we mean a single measure that characterizes a property. Since previous literature has reported that different categories within a language can differ in their patterns of transfer, and that shifts in individual categories may not be identical to the patterns of system-wide shift, we cannot have a single indivisibe ("atomic") measure such as "extent of transfer in L1"-- we must view the overall transfer as resulting from shifts constrained by different factors (cognitive, linguistic). To do this, we need to examine transfer in a wider range of sound contrasts.

10.     Page 19, section 1.7. Not sure this list merits its own section. I believe this should be incorporated in the previous section. I also believe the question might need to be revised. You do not only check if vowel categories are modified in one setting relative to the other setting. You also check the direction and size (or extent) of the modification. Should you have a main research question and some follow-up, secondary questions focused on directionality and extent-of-change?
18.     Page 46, line 9. “Our third research question Paradigm concerned…” I am a bit confused. I understood “Paradigm” to be a hypothesis, not a research question. Please, check section 1.7 and decide if Asymmetry and Paradigm and research questions or hypotheses. If they are questions, what are the hypotheses in this study? Also, when referring to Paradigm and Asymmetry in the Discussion section, please notice that, in most cases, surrounding commas are needed: “Our third research question, Paradigm, concerned…”

AR: We have modified our Research Questions and Hypotheses section on the basis of these comments (pg. ...).

AR: In this paragraph, we suggest that (i) examining multiple acoustic features within a single study would allow us to compare predictions made by different models for phonetic transfer; (ii) given its communicative relevance (as demonstrated by sociophonetic literature), communication-centered approaches might be more appropriate than inhibition-based ones. For the same reason, data from native multilingual populations is valuable, since these settings provide obvious communicative motivations for transfer. We have revised the wording in the paragraph, and hope it comes across more clearly now.

@@

14.     Page 37, lines 46-51. Please, revise sentence top include figure numbers, although you will hopefully delete the whole section.

15.     Page 42, line 26-27. “it is a potential”. What does “it” refer to? Please, spell out, e.g.: “, differences reading proficiency are a potential…”

17.     Page 44, line 11: “in a languages” > “in a language” or “in languages”

AR: Thanks for pointing these out! We have revised the relevant parts of the manuscript.

@@

Congratulations on a fine study and thank you for the opportunity to learn about this study before it was published. I hope some of my comments are found to be useful.

AR: Thank you for the detailed and constructive feedback! We hope we have been able to address all the concerns.

### 2022-10-10 draft_BLC/BLC_AM-ID_manuscript.tex
added reviewer response

Word count (29.7.2022): 13388 -- remove 4388

@@

Bengali (also, Bangla) is an Indo-Aryan language primarily spoken in India and Bangladesh. In India, more than 97 million people speak Bengali as a first language (Census of India, 2011), mostly in the state of West Bengal. A majority of this population also speaks additional languages.

Indian English (IE) refers to the variety of English that has developed in the Indian subcontinent. In India, it is spoken as an L2 by 129 million people (Census of India, 2011).  English is one of the two official languages, used in education, law, media, as a lingua franca mainly for an educated elite in most metropolitan regions, and carries a high prestige value \citep{pandey201517, kachru1981english, tollefson2014language, kachru1983indianization}.

@@

The socio-linguistic facts about English in India also suggest that mixed-language processing of English and an L1 an ecologically valid paradigm to understand short-term phonetic interaction.

@@

The vowel system of IE contains the monophthongs \nt{I, i, E, e, \ae, @/2/3:, a:, O, o, U, u} \citep{wells1982accents, masica1972sound}. The vowel chart in figure \ref{vowels_ie} shows their distribution in the F1$\times$F2 formant space. 
A single mid-central vowel corresponds to the categories \nt{2,@,3:}, which are treated as distinct in many native varieties of English \citep{nihalani1979indian,wells1982accents,hickey2005legacies,bansal1969intelligibility}. Since the English items used in this study are traditionally transcribed with \nt{2}, we use this symbol to indicate the mid-central vowel throughout.

@@

Existing research has distinguished changes in category \textsc{representations} due to the acquisition of multiple sound systems (cf. Speech Learning Model \citep{flege1995second,flege2007language}; Perceptual Assimilation Model-L2 \citep{best2007nonnative}), from interaction during accessing, processing, or articulation of these categories (cf. transfer vs. interference \citep{grosjean2012attempt}; competence vs. performance interference \citep{paradis1993linguistic}). 
Given their transient nature, dynamic changes in production during mixed-language use are generally attributed to the latter, e.g. online processing costs \citep[][VOT]{olson2013bilingual,tsui2019impact,vsimavckova2015immediate}, language mode \citep[][vowel quality]{simonet2014phonetic}, context-awareness \citep[][phonological variables]{khattab2013phonetic}. What triggers this interaction? \cite{olson2016role} argues that cross-language phonetic effects could have two potential sources: (i) the local point of switch, and (ii) global co-activation of two languages: \textsc{bilingual language mode} \citep{grosjean1998studying}.

@@

Since participants of this study are in an environment that largely contains mixed-language input and multilingual interlocutors, we expect that phonetic transfer during everyday language use, if any, takes place in a bilingual mode \citep[c.f.][]{grosjean1998studying}. Thus, we elicited all utterances in the same bilingual test block. Any observed differences in this paradigm would arguably result from interaction during online processing of sounds while switching between languages.

@@

Longitudinal studies and between-subject comparisons of speakers with different durations of L2 exposure suggest that interaction between bilingual sound systems is dynamic \citep[e.g.][]{bohn1992production}. Moreover, changes due to transfer are not necessarily unidirectional or irreversible. \cite[][VOT]{sancier1997gestural} first demonstrated that spending 2-5 months in an L1 or L2 environment causes productions in both languages to ``drift" towards the ambient language, showing that cross-language interaction can be triggered in an order of months, and also reversed within a similar time range. A more recent study by \cite[][VOT]{tobin2017phonetic} reported comparable effects in an even shorter duration (2-4 weeks). Based on a short-term longitudinal study, \cite[][VOT]{chang2012rapid} demonstrates that over the course of the first five weeks of learning an L2, there is a gradual convergence of L1 towards L2. For some sounds and features (though not others, cf. sec.\ref{asymmetry between sounds}), transfer was additive over time. How long these effects last in the absence of regular L2 input was not tested.

@@

\subsubsection*{L1 vs. L2 status} 
Flege's Speech Learning Model (SLM) \citeyearpar{flege1995second,flege2007language} proposes that the sound categories of a bilingual speaker exist in a common phonological space, and therefore in principle, both L1 and L2 categories can influence one another. Thus, the patterns of cross-language transfer depend on how L2 phonemes are mapped in relation to the existing L1 categories. In non-switched production, this is evidenced through a L1 influence on non-native contrasts that are perceptually linked to an existing L1 contrast (cf. \textsc{equivalence classification}: \cite{flege1984limits,flege1987production}), and the observation that both L1 and L2 sound systems of bilinguals differ from those of corresponding monolingual speakers \citep[e.g.][vowel quality]{guion2003vowel}.

@@

In the present study, we focus on two vowel categories in Indian English: the mid-central vowel \nt{2} and the low front vowel \nt{\ae}. There are two plausible sources of asymmetry between these:
\begin{enumerate}
	\item Position in the IE vowel space: compared to \nt{\ae}, \nt{2} exists in a part of the vowel space that has a lower vowel density. This affords a greater latitude for movement without risking the loss of a contrast, particularly in the vowel height (F1) dimension. Thus, considering purely phonological constraints on IE, we expect that a greater degree of shift is possible in \nt{2} compared to \nt{\ae}. Any movement in \nt{\ae} is expected to be primarily in the backness (F2) dimension.
	\item Target of transfer: findings from existing research on cross-language phonetic interaction suggest that changes in production during mixed-language use are not random, but rather targeted with respect to categories in the other language. Thus, another source of asymmetry is the fact that the category \nt{2} is absent in Bengali, whereas \nt{\ae} is a common category across both languages (cf. sec.\ref{vowel systems}).
\end{enumerate}

@@

This section motivates a secondary aim of this study: to verify if the extent and patterns of phonetic interaction in bilingual participants are independently mediated by the choice of experimental paradigm. Studies of phonetic transfer during mixed-language use have largely employed spontaneous or scripted code-switching (CS) to elicit mixed productions. CS is defined as the use of multiple languages within a single utterance, i.e. in connected speech \citep{myers1993dueling}. Other paradigms which have been used include cued picture-naming, delayed repetition, interpreting across languages, reading word-lists etc. \cite{olson2013bilingual} argues that code-switching introduces discursive factors such as discourse planning and pragmatic considerations, which ultimately reflect patterns of language practice, rather than cognitive behavior. This finds support from studies which report transfer effects \textsc{before} the switch point \citep{bullock2009trying}, or other evidence for planning such as hyperarticulation \citep{muldner2019phonetics} and interlocutor-awareness in spontaneous CS \citep{khattab2013phonetic}. To avoid such confounds, \cite{olson2013bilingual} endorses cued picture-naming (where pictures are named in rapid succession in either language, based on the given language cue) as an alternative paradigm to isolate purely phonetic effects in controlled setups. This observation has important implications for how the existing literature on short-term phonetic interaction is interpreted: if the discursive properties of a CS paradigm independently affect the phonetic outcome of mixed-language use, this calls to question the comparability of results across studies using these different paradigms. Because the reported outcomes of short-term phonetic interaction vary so greatly across studies (cf. section \ref{asymmetries}), generalizations about the mechanisms that underlie these patterns often rely crucially on the assumption of such comparability. Therefore, identifying paradigm-specific differences is particularly relevant for this line of research.

@@

\subsection*{Research questions and hypotheses}\label{questions_and_hypotheses}
\begin{itemize}
	\item \textbf{RQ1---L2 vowels}: Are L2 vowel categories of proficient bilingual speakers affected by dynamic phonetic transfer during mixed-language production?
	\item \textbf{H1---L2 vowels}: We expect to see differences in the spectral properties of L2 vowels in switched vs non-switched tokens
	\item \textbf{RQ2---Asymmetry}: Do patterns of transfer differ between the two vowel categories?
	\item \textbf{H2---Asymmetry}: Since existing literature demonstrates targeted and phonologically-constrained shifts, we expect transfer effects to be category-specific. Based on the position of IE vowels compared to those of Bengali vowels from \cite{shruticorpus}, we expect \nt{2} to be lowered and fronted in the mixed condition. \nt{\ae} is expected to be fronted. We hypothesize that the extent of shift in \nt{\ae} to be lesser than \nt{2}, as it is in a part of the IE vowel space that has greater vowel density.
	\item \textbf{RQ3---Paradigm}: Is there an observable difference between the extent of transfer in a cued picture-naming vs. a code-switching paradigm? 
	\item \textbf{H3---Paradigm}: While researchers have discussed code-switching as a source of discursive factors likely to affect transfer behavior, this has not been systematically studied. Existing literature does not allow us to predict precisely what form this effect might take (c.f. discussion in Section \ref{paradigms}). Thus, this question is exploratory, and meant to serve as a starting point for more within-study comparisons. 
\end{itemize}

@@

As noted in section \ref{bengali_english_in_india}, the population is multilingual and linguistically diverse. The university campus comprises of students from different parts of the country and world, with varying L1s and different varieties of English. Our aim in screening participants was to minimize between-participant random variability, while maintaining ecological validity considering the actual day-to-day settings in which transfer behaviors likely take place. We chose two metrics to do this: (i) where the participant grew up; and (ii) amount of formal education in Bengali. The first ensures that the variety of Bengali that individuals are exposed to in their surroundings is similar across participants. This would mean that the target representations for Bengali vowels would be expected to be similar. This was especially important because we were not eliciting unilingual L1 Bengali data in this study, and were using an existing data source to estimate the position of Bengali vowels in the formant space. Based on previous literature (c.f. section \ref{asymmetry between sounds}), we expect that L2 vowels shift towards related L1 categories during mixed language utterances. Thus, we wanted to ensure that any between-participant asymmetries in the pattern of shift reflected individual differences in transfer, rather than different acoustic targets. Another reason to control for location was that it is a fairly reliable predictor for language(s) of education(see Supplementary materials for details).

@@

A total of 28 bilingual speakers of Bengali and English (18 female) responded to the recruitment call. Using the inclusion criteria reported above, 10 volunteers (5 female, 5 male, age range 19 to 28) were invited to participate in a speech production study, and were compensated for their time. After completion of the study, we followed up with a detailed language background survey at a later date to learn more about the language profiles of the participants. This survey was a modified and consolidated version of three popular bilingualism profiling tools: the Bilingual Language Profile (BLP) \citep{blp}, the Language Experience and Proficiency Questionnaire (LEAP-Q) \citep{leap-q} which incorporates language history and self-assessed proficiency, and the Bilingual Switching Questionnaire (BSWQ) \citep{language_switching_questionnaire}. Together, these incorporate questions about language history, usage, attitudes, and switching experience. Table \ref{table_participants} summarizes some key biographical data from the survey. \alert{The complete questionnaire, along with responses, is available in the ``Data" component of the OSF repository for this project (c.f. section \ref{data_availability}). A more detailed discussion of the linguistic situation and language profiles is included as supplementary material.--footnote?}

Given our inclusion criteria, we were able to recruit 10 participants for this study. Although existing studies in the field have used comparable participant numbers, we acknowledge that this is a small sample size. As detailed above, even after a recruitment call that already controlled for language background and education level, we were able to include only 10 out of 28 volunteers (35.7\%) in the final study. The criteria used here establish only a minimal level of between-participant similarity in exposure, dominance, habits, and proficiency, such as is extremely common in most comparable studies. This demonstrates the degree of linguistic variability in the present population. Norms for experimental research on bilingualism have largely been established in a Western context. In settings where linguistic variability at the population level is much greater than these canonically studied groups, establishing comparable levels of experimental control requires significantly greater resources. Given that small sample sizes are likely to have low statistical power, replication with larger sample sizes are important to confirm the observed effects, and provide much-needed data diversity. A more detailed discussion of power analyses in bilingualism research is included as supplementary material on OSF (c.f. section \ref{data_availability}). We hope that this can facilitate more conversation around establishing norms for power and sample size in our field.

\subsection{Stimuli}
Twenty monosyllabic English words: ten containing the vowel \nt{2} and ten containing \nt{\ae}, were selected as target words. To minimize coarticulatory effects on the vowel, the pre-vocalic consonants were all voiced plosives (\nt{b} or \nt{d}). 10 monosyllabic English words that do not contain the target vowel served as filler items. We avoided English words that are already lexicalized loanwords in Bengali. The complete list of target words can be found in the ``Materials" component of the OSF repository for this project (c.f. section \ref{data_availability}).

@@

\begin{enumerate}[]
	\item Cued picture-naming: Participants were presented with slides in the following sequence:
	\begin{itemize}
		\item Language cue: a word in either English or Bengali orthography
		\item A picture displayed for 50 ms; named by the participant in either English or Bengali, depending on the language cue. The picture of a fish was used in all trials, prompting the participant to say either \nt{fIS} or \nt{ma: \textteshlig \super{h}}, depending on the language cue
		\item Target: an English word; read out by the participant
		\item Distracter math problem
	\end{itemize}
	Since all the target words were in English, the language cue was used to manipulate the preceding word, giving target words in non-switched and switched (mixed) utterances. The orthographies of English and Bengali are visually distinct, and were therefore used to cue language.

@@

Carrier sentences were chosen such that the sound preceding the target word was uniform across all sentences (the final sound in \textit{yellow} is produced as [o] in Indian English), to minimize any possible differences due to coarticulatory effects on the target vowel. To ensure identical word order in both languages, we chose a specifier-noun construction, as this has a parallel structure in Bengali and English. This gave a uniform prosodic context for the target word across conditions. While presenting the mixed-language utterances, we used mixed orthography. \\

>>>>

Word count (7.10.2022): 9650

@@

Bengali is an Indo-Aryan language primarily spoken in India and Bangladesh. In India, more than 97 million people speak Bengali as a first language (Census of India, 2011), mostly in the state of West Bengal. A majority of this population also speaks additional languages.

Indian English (IE) refers to the variety of English spoken in the Indian subcontinent. It is spoken as an L2 by 129 million people (Census of India, 2011).  English is one of two official languages, used in education, law, media, as a lingua franca mainly by an educated elite in most metropolitan regions, and carries high prestige value \citep{pandey201517, kachru1981english, tollefson2014language, kachru1983indianization}.

@@

The socio-linguistic facts around English in India also suggest that mixed-language processing of English and an L1 is an ecologically valid paradigm to understand short-term phonetic interaction.

@@

The vowel system of IE contains the monophthongs \nt{I, i, E, e, \ae, @/2/3:, a:, O, o, U, u} \citep{wells1982accents, masica1972sound}\footnote{\nt{@, 2, and 3:} are treated as a single phoneme in many descriptions of IE; the slashes between the symbols indicate possible notations for the same category}. The vowel chart in figure \ref{vowels_ie} shows their distribution in the F1$\times$F2 formant space. 
A single mid-central vowel corresponds to the categories \nt{2,@,3:}, which are treated as distinct in many native varieties of English \citep{nihalani1979indian,wells1982accents,hickey2005legacies,bansal1969intelligibility}. Since the English items used in this study are traditionally transcribed with \nt{2}, we use this symbol to indicate the mid-central vowel throughout.

@@

Existing research has distinguished changes in category \textsc{representations} due to the acquisition of multiple sound systems (Speech Learning Model \citep{flege1995second,flege2007language}; Perceptual Assimilation Model-L2 \citep{best2007nonnative}), from interaction during accessing, processing, or articulation of these categories (transfer vs. interference \citep{grosjean2012attempt}; competence vs. performance interference \citep{paradis1993linguistic}). 
Given their transient nature, dynamic changes in production during mixed-language use are generally attributed to the latter, e.g. online processing costs \citep[][VOT]{olson2013bilingual,tsui2019impact,vsimavckova2015immediate}, language mode \citep[][vowel quality]{simonet2014phonetic}, context-awareness \citep[][phonological variables]{khattab2013phonetic}. What triggers this interaction? \cite{olson2016role} argues that cross-language phonetic effects could have two potential sources: (i) the local point of switch, and (ii) global co-activation of two languages: \textsc{bilingual language mode} \citep{grosjean1998studying}.

@@

Since participants of this study are in an environment that largely contains mixed-language input and multilingual interlocutors, we expect that phonetic transfer during everyday language use, if any, takes place in a bilingual mode \citep[see][]{grosjean1998studying}. Thus, we elicited all utterances in the same bilingual test block. Any observed differences in this paradigm would arguably result from interaction during online processing of sounds while switching between languages.

@@

Longitudinal studies and between-subject comparisons of speakers with different durations of L2 exposure suggest that interaction between bilingual sound systems is dynamic \citep[e.g.][]{bohn1992production}. Moreover, changes due to transfer are not necessarily unidirectional or irreversible. \cite[][VOT]{sancier1997gestural} first demonstrated that spending 2-5 months in an L1 or L2 environment causes productions in both languages to ``drift" towards the ambient language, showing that cross-language interaction can be triggered in an order of months, and also reversed within a similar time range. A more recent study by \cite[][VOT]{tobin2017phonetic} reported comparable effects in an even shorter duration (2-4 weeks). Based on a short-term longitudinal study, \cite[][VOT]{chang2012rapid} demonstrates that over the course of the first five weeks of learning an L2, there is a gradual convergence of L1 towards L2. For some sounds and features (though not others, see sec.\ref{asymmetry between sounds}), transfer was additive over time. How long these effects last in the absence of regular L2 input was not tested.

@@

\subsubsection*{L1 vs. L2 status} 
Flege's Speech Learning Model (SLM) \citeyearpar{flege1995second,flege2007language} proposes that the sound categories of a bilingual speaker exist in a common phonological space, and therefore in principle, both L1 and L2 categories can influence one another. Thus, the patterns of cross-language transfer depend on how L2 phonemes are mapped in relation to the existing L1 categories. In non-switched production, this is evidenced through pervasive L1 influence on non-native contrasts that are perceptually linked to an existing L1 contrast (cf. \textsc{equivalence classification}: \cite{flege1984limits,flege1987production}), and the observation that both L1 and L2 sound systems of bilinguals differ from those of corresponding monolingual speakers \citep[e.g.][vowel quality]{guion2003vowel}.

@@

In the present study, we focus on two vowel categories in Indian English: the mid-central vowel \nt{2} and the low front vowel \nt{\ae}. There are two plausible sources of asymmetry between these:
\begin{enumerate}
	\item Position in the IE vowel space: compared to \nt{\ae}, \nt{2} exists in a part of the vowel space that has a lower vowel density. This affords a greater latitude for movement without risking the loss of a contrast, particularly in the vowel height (F1) dimension. Thus, considering purely phonological constraints on IE, we expect that a greater degree of shift is possible in \nt{2} compared to \nt{\ae}. Any movement in \nt{\ae} is expected to be primarily in the backness (F2) dimension.
	\item Target of transfer: findings from existing research on cross-language phonetic interaction suggest that changes in production during mixed-language use are not random, but rather targeted with respect to categories in the other language. Thus, another source of asymmetry is the fact that the category \nt{2} is absent in Bengali, whereas \nt{\ae} is a common category across both languages (see sec.\ref{vowel systems}).
\end{enumerate}

@@

This section motivates a secondary aim of this study: to verify if the extent and patterns of phonetic interaction in bilingual participants are independently mediated by the choice of experimental paradigm. Studies of phonetic transfer during mixed-language use have largely employed spontaneous or scripted code-switching (CS) to elicit mixed productions. CS is defined as the use of multiple languages within a single utterance, i.e. in connected speech \citep{myers1993dueling}. Other paradigms which have been used include cued picture-naming, delayed repetition, interpreting across languages, reading word-lists etc. \cite{olson2013bilingual} argues that code-switching introduces discursive factors such as discourse planning and pragmatic considerations, which ultimately reflect patterns of language practice, rather than cognitive behavior. This finds support from studies which report transfer effects \textsc{before} the switch point \citep{bullock2009trying}, or other evidence for planning such as hyperarticulation \citep{muldner2019phonetics} and interlocutor-awareness in spontaneous CS \citep{khattab2013phonetic}. To avoid such confounds, \cite{olson2013bilingual} endorses cued picture-naming (where pictures are named in rapid succession in either language, based on the given language cue) as an alternative paradigm to isolate purely phonetic effects in controlled setups. This observation has important implications for how the existing literature on short-term phonetic interaction is interpreted: if the discursive properties of a CS paradigm independently affect the phonetic outcome of mixed-language use, this calls to question the comparability of results across studies using these different paradigms. Because the reported outcomes of short-term phonetic interaction vary so greatly across studies (see section \ref{asymmetries}), generalizations about the mechanisms that underlie these patterns often rely crucially on the assumption of such comparability. Therefore, identifying paradigm-specific differences is particularly relevant for this line of research.

@@

\subsection*{Research questions and hypotheses}\label{questions_and_hypotheses}
\begin{itemize}
	\item \textbf{RQ1---L2 vowels}: Are L2 vowel categories of proficient bilingual speakers affected by dynamic phonetic transfer during mixed-language production?
	\item \textbf{H1---L2 vowels}: We expect to see differences in the spectral properties of L2 vowels in switched vs non-switched tokens
	\item \textbf{RQ2---Asymmetry}: Do patterns of transfer differ between the two vowel categories?
	\item \textbf{H2---Asymmetry}: Since existing literature demonstrates targeted and phonologically-constrained shifts, we expect transfer effects to be category-specific. Based on the position of IE vowels compared to those of Bengali vowels from \cite{shruticorpus}, we expect \nt{2} to be lowered and fronted in the mixed condition. \nt{\ae} is expected to be fronted. We hypothesize that the extent of shift in \nt{\ae} to be lesser than \nt{2}, as it is in a part of the IE vowel space that has greater vowel density.
	\item \textbf{RQ3---Paradigm}: Is there an observable difference between the extent of transfer in a cued picture-naming vs. a code-switching paradigm? 
	\item \textbf{H3---Paradigm}: While researchers have discussed code-switching as a source of discursive factors likely to affect transfer behavior, this has not been systematically studied. Existing literature does not allow us to predict precisely what form this effect might take (see discussion in Section \ref{paradigms}). Thus, this question is exploratory, and meant to serve as a starting point for more within-study comparisons. 
\end{itemize}

@@

As noted in section \ref{bengali_english_in_india}, the population is multilingual and linguistically diverse. The university campus comprises of students from different parts of the country and world, with varying L1s and different varieties of English. Our aim in screening participants was to minimize between-participant random variability, while maintaining ecological validity considering the actual day-to-day settings in which transfer behaviors likely take place. We chose two metrics to do this: (i) where the participant grew up; and (ii) amount of formal education in Bengali. The first ensures that the variety of Bengali that individuals are exposed to in their surroundings is similar across participants. This would mean that the target representations for Bengali vowels would be expected to be similar. This was especially important because we were not eliciting unilingual L1 Bengali data in this study, and were using an existing data source to estimate the position of Bengali vowels in the formant space. Based on previous literature (see section \ref{asymmetry between sounds}), we expect that L2 vowels shift towards related L1 categories during mixed language utterances. Thus, we wanted to ensure that any between-participant asymmetries in the pattern of shift reflected individual differences in transfer, rather than different acoustic targets. Another reason to control for location was that it is a fairly reliable predictor for language(s) of education(see Supplementary materials for details).

@@

A total of 28 bilingual speakers of Bengali and English (18 female) responded to the recruitment call. Using the inclusion criteria reported above, 10 volunteers (5 female, 5 male, age range 19 to 28) were invited to participate in a speech production study, and were compensated for their time. After completion of the study, we followed up with a detailed language background survey at a later date to learn more about the language profiles of the participants. This survey was a modified and consolidated version of three popular bilingualism profiling tools: the Bilingual Language Profile (BLP) \citep{blp}, the Language Experience and Proficiency Questionnaire (LEAP-Q) \citep{leap-q} which incorporates language history and self-assessed proficiency, and the Bilingual Switching Questionnaire (BSWQ) \citep{language_switching_questionnaire}. Together, these incorporate questions about language history, usage, attitudes, and switching experience. Table \ref{table_participants} summarizes some key biographical data from the survey. \alert{The complete questionnaire, along with responses, is available in the ``Data" component of the OSF repository for this project (see section \ref{data_availability}). A more detailed discussion of the linguistic situation and language profiles is included as supplementary material.--footnote?}

Given our inclusion criteria, we were able to recruit 10 participants for this study. Although existing studies in the field have used comparable participant numbers, we acknowledge that this is a small sample size. As detailed above, even after a recruitment call that already controlled for language background and education level, we were able to include only 10 out of 28 volunteers (35.7\%) in the final study. The criteria used here establish only a minimal level of between-participant similarity in exposure, dominance, habits, and proficiency, such as is extremely common in most comparable studies. This demonstrates the degree of linguistic variability in the present population. Norms for experimental research on bilingualism have largely been established in a Western context. In settings where linguistic variability at the population level is much greater than these canonically studied groups, establishing comparable levels of experimental control requires significantly greater resources. Given that small sample sizes are likely to have low statistical power, replication with larger sample sizes are important to confirm the observed effects, and provide much-needed data diversity. A more detailed discussion of power analyses in bilingualism research is included as supplementary material on OSF (see section \ref{data_availability}). We hope that this can facilitate more conversation around establishing norms for power and sample size in our field.

\subsection{Stimuli}
Twenty monosyllabic English words: ten containing the vowel \nt{2} and ten containing \nt{\ae}, were selected as target words. To minimize coarticulatory effects on the vowel, the pre-vocalic consonants were all voiced plosives (\nt{b} or \nt{d}). 10 monosyllabic English words that do not contain the target vowel served as filler items. We avoided English words that are already lexicalized loanwords in Bengali. The complete list of target words can be found in the ``Materials" component of the OSF repository for this project (see section \ref{data_availability}).

@@

\begin{enumerate}[]
	\item Cued picture-naming: Participants were presented with slides in the following sequence (see fig. \ref{figure_picture_task}):
	\begin{itemize}
		\item Language cue: a word in either English or Bengali orthography, not read aloud
		\item A picture displayed for 50 ms; named aloud by the participant in either English or Bengali, depending on the language cue. The picture of a fish was used in all trials, prompting the participant to say either \nt{fIS} (English) or \nt{ma: \textteshlig \super{h}} (Bengali)
		\item Target: an English word; read aloud by the participant
		\item Distracter math problem
	\end{itemize}
	The language cue was used to manipulate the word preceding the target, giving target words in non-switched and switched (mixed) utterances.

@@

Carrier sentences were chosen such that the sound preceding the target word was uniform across all sentences (the final sound in \textit{yellow} is produced as [o] in Indian English), to minimize differences due to coarticulatory effects on the target vowel. To ensure identical word order in both languages, we chose a specifier-noun construction, as this has a parallel structure in Bengali and English. This gave a uniform prosodic context for the target word across conditions. While presenting the mixed-language utterances, we used mixed orthography. \\

<insert fig. \ref{figure_picture_task} about here>

\end{enumerate}
\subsection{Acoustic analysis}\label{analysis}
The first iteration of each task was treated as a trial, and excluded from further analyses. An additional 67 tokens (2.79\%) were excluded due to errors in presentation/production, giving 2333 target word tokens (\nt{2}: 1137, \nt{\ae}: 1196) for analysis. Recordings were manually segmented and annotated in Praat \citep{boersma2016praat}, and the first and second formant frequencies (F1 and F2) of the target vowels \nt{2, \ae} were measured at five points: in 10\% increments starting at 5\% into the vowel. This gave measures of F1 and F2 at 5\%, 15\%, 25\%, 35\% and 45\% into the vowel, allowing us to observe any dynamic changes in the extent of transfer over the course of producing the vowel. Individual vocal tract differences affect formant frequencies across speakers. To reduce between-speaker differences, we used z-score normalization (using the PhonR package \citep{phonR} in R \citep{r}, representing each speaker's formant frequencies as the scaled distance from their own mean). \footnote{Since our experimental treatments were between-subject and statistical analyses use mixed-effects regression, this step would not affect the results, as pointed out by a reviewer. This was done to allow for more intuitive values and visualizations, rater than statistical necessity. The code for parallel statistical analysis using non-z-score-normalized values can be found in the OSF repository.} Plotting these vowels shows a difference between male and female speakers, such that male speakers have overall lower formant frequencies (figure \ref{gender_before}). %Speaker normalization using the z-score method does not get rid of this difference. 
Since gender-related differences in formant frequencies are not of interest in this study, we normalized individual values by gender (representing each value as a scaled distance from the group mean; figure \ref{gender_after}). Statistical analyses were performed on these normalized values.
